Interacting with Modal Logics \\ in the Coq Proof Assistant

### 2014-12-12 Papers/InProgress/ModalLogicCoq/ModalCoq.tex
Merge conflicting hunks

In order to illustrate the use of the implemented embedding, we show
here the formalization of Scott's version \cite{ScottNotes} of G\"odel's
ontological argument for God's existence (in Section \ref{sec:proof}).
This proof was chosen mainly for two reasons. Firstly, it requires not
only modal operators, but also higher-order quantification. Therefore,
it is beyond the reach of specialized propositional and first-order
(modal) theorem provers. Secondly, this argument addresses an ancient
problem in Philosophy and Metaphysics, which has nevertheless received
a lot of attention in the last 15 years, because of the discovery of
the modal collapse \cite{sobel,sobel2004logic}. This proof lies in the center of a
vast and largely unexplored application domain for automated and
interactive theorem provers.

@@

Rules for $\alpha\beta\eta$-equality and axioms (or rules) for extensionality are omitted here since they are not important for the rest of the paper. For a full, sound and Henkin-complete, classical higher-order ND calculus see \cite{HOLND}.
\end{small}

@@

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they give rise to the inference
rules\footnote{The ND calculus with the rules from
Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
complete relatively to the calculus of Figure
\ref{fig:NaturalDeduction} extended with a necessitation rule and the
modal axiom K \cite{ToDo:UnpublishedPaperInGithub}. Starting from a sound and Henkin-complete ND calculus for classical higher-order logic (cf. Figure~\ref{fig:NaturalDeduction}), the additional
modal rules in Figure \ref{fig:NDK} make it sound and Henkin-complete
for modal higher-order logic \textbf{K}.} shown in Figure
\ref{fig:NDK}. The labels that name boxes in those inference rules are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\bibliographystyle{plain}
%\bibliography{Bibliography}
\begin{thebibliography}{10}

\bibitem{Adams}
R.M.~Adams.
\newblock Introductory note to *1970.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1995.

\bibitem{AndersonGettings}
A.C.~Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

\bibitem{HOLND}
C.~Benzm{\"u}ller, C.~Brown, and M.~Kohlhase.
\newblock Higher-Order Semantics and Extensionality. 
\newblock {\em Journal of Symbolic Logic} 69(4):1027-1088, 2004.

\bibitem{JensOtten}
C.~Benzm{\"u}ller, J. Otten, and Th. Raths.
\newblock Implementing and Evaluating Provers for First-order Modal
Logics. 
\newblock In {\em ECAI 2012}, pages 163-168. IOS Press, 2012.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C.~Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications.

\bibitem{J23}
C.~Benzm{\"u}ller and L.C.~Paulson.
\newblock Quantified multimodal logics in simple type theory.
\newblock {\em Logica Universalis (Special Issue on Multimodal Logics)},
  7(1):7--20, 2013.

\bibitem{LEO-II}
C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke.
\newblock {LEO-II} - a cooperative automatic theorem prover for higher-order
  logic.
\newblock In {\em Proc. of IJCAR 2008}, volume 5195 of {\em LNAI}, pages
  162--170. Springer, 2008.

\bibitem{arXiv}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock Formalization, Mechanization and Automation of G{\"o}del's Proof of God's Existence.
\newblock {\em {arXiv}}:1308.4526, 2013.

\bibitem{AFP}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock G{\"o}del's God in Isabelle/HOL. 
\newblock {\em Archive of Formal Proofs}, 2013.

\bibitem{Coq}
Y.~Bertot and P.~Casteran.
\newblock {\em {Interactive Theorem Proving and Program Development}}.
\newblock Springer, 2004.

\bibitem{ModalLogic}
P.~Blackburn, J.v.~Benthem, and F.~Wolter (eds.), \newblock {\em {Handbook of Modal Logic}}. Elsevier, 2006.

\bibitem{Sledgehammer}
J.C.~Blanchette, S.~B\"ohme, and L.C.~Paulson.
\newblock Extending {Sledgehammer} with {SMT} solvers.
\newblock {\em Journal of Automated Reasoning}, 51(1):109--128, 2013.

\bibitem{Nitpick}
J.C.~Blanchette and T.~Nipkow.
\newblock Nitpick: A counterexample generator for higher-order logic based on a
  relational model finder.
\newblock In {\em Proc. of ITP 2010}, no.~6172 in LNCS, pages 131--146.
  Springer, 2010.

\bibitem{Satallax}
C.E.~Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
  Springer, 2012.

\bibitem{CIC}
Th.~Coquand and G.~Huet. 
\newblock The Calculus of Constructions. 
\newblock {\em Information and Computation} 76 (2-3). 1988.

\bibitem{ContemporaryBibliography}
R.~Corazzon.
\newblock Contemporary bibligraphy on the ontological proof
  (\url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}).

\bibitem{Fitting}
M.~Fitting.
\newblock {\em {Types, Tableaux and G\"odel's God}}.
\newblock Kluver Academic Press, 2002.

\bibitem{Labels}
D.M.~Gabbay. {\em Labelled Deductive Systems}. Clarendon Press, 1996.

\bibitem{Goedel1970}
K.~G\"odel.
\newblock Ontological proof.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1970.

\bibitem{GoedelNotes}
K.~G\"odel.
\newblock {\em Appendix A. Notes in Kurt G\"odel's Hand}, pages 144--145.
\newblock In  \cite{sobel2004logic}, 2004.

\bibitem{HOLLight}
J.~Harrison.
\newblock HOL Light: An overview.
\newblock In {\em Proceedings of TPHOLs 2009}. Number 5674 in LNCS, pp. 60-66. Springer, 2009.

\bibitem{Hazen}
A.P.~Hazen.
\newblock On G\"odel's ontological proof.
\newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

\bibitem{Hurd03first-orderproof}
J.~Hurd.
\newblock First-order proof tactics in higher-order logic theorem provers.
\newblock In {\em Design and Application of Strategies/Tactics in Higher Order
  Logics, NASA Tech. Rep. NASA/CP-2003-212448}, 2003.

\bibitem{Isabelle}
T.~Nipkow, L.C. Paulson, and M.~Wenzel.
\newblock {\em {Isabelle/HOL: A Proof Assistant for Higher-Order Logic}}.
\newblock Number 2283 in LNCS. Springer, 2002.

\bibitem{Zalta}
P.E.~Oppenheimera and E.N.~Zalta.
\newblock A Computationally-Discovered Simplification of the Ontological Argument.
\newblock {\em Australasian Journal of Philosophy}, 89(2):333-349, 2011.

\bibitem{Rushby}
J.~Rushby.
\newblock The Ontological Argument in PVS. 
\newblock In {\em Proc.~of CAV Workshop ``Fun With Formal Methods''}, St. Petersburg, Russia, 2013.

\bibitem{ScottNotes}
D.~Scott.
\newblock {\em Appendix B. Notes in Dana Scott's Hand}, pages 145--146.
\newblock In  \cite{sobel2004logic}, 2004.

\bibitem{sobel}
J.H.~Sobel. 
\newblock G\"odel's ontological proof. 
\newblock {\em In On Being and Saying. Essays for Richard Cartwright}, pages 241-261. MIT Press, 1987.

\bibitem{sobel2004logic}
J.H.~Sobel.
\newblock {\em Logic and Theism: Arguments for and Against Beliefs in God}.
\newblock Cambridge U. Press, 2004.

\bibitem{J22}
G.~Sutcliffe and C.~Benzm{\"u}ller.
\newblock Automated reasoning in higher-order logic using the {TPTP THF}
  infrastructure.
\newblock {\em Journal of Formalized Reasoning}, 3(1):1--27, 2010.

\bibitem{FormalTheologyRepository}
B.~Woltzenlogel Paleo and C.~Benzm\"uller.
\newblock Formal theology repository
  (\url{https://github.com/FormalTheology/GoedelGod}).
\end{thebibliography}

>>>>

In order to illustrate the use of the implemented embedding, we show
here the formalization of Scott's version \cite{ScottNotes} of G\"odel's
ontological argument for God's existence (in Section \ref{sec:proof}).
This proof was chosen mainly for two reasons. Firstly, it requires not
only modal operators, but also higher-order quantification. Therefore,
it is beyond the reach of specialized propositional and first-order
(modal) theorem provers. Secondly, this argument addresses an ancient
problem in Philosophy and Metaphysics, which has nevertheless received
a lot of attention in the last 15 years, because of the discovery of
the modal collapse \cite{Sobel,sobel2004logic}. This proof lies in the center of a
vast and largely unexplored application domain for automated and
interactive theorem provers.

@@

Rules for $\alpha\beta\eta$-equality and axioms (or rules) for extensionality are omitted here since they are not important for the rest of the paper. For a full, sound and Henkin-complete, classical higher-order ND calculus, see \cite{HOLND}.
\end{small}

@@

\noindent If the new modal tactics above are regarded from a natural
deduction point of view, they give rise to the inference
rules\footnote{The ND calculus with the rules from
Figures \ref{fig:NaturalDeduction} and \ref{fig:NDK} is sound and
complete relatively to the calculus of Figure
\ref{fig:NaturalDeduction} extended with a necessitation rule and the
modal axiom K \cite{ModalND}. Starting from a sound and Henkin-complete ND calculus for classical higher-order logic (cf. Figure~\ref{fig:NaturalDeduction}), the additional
modal rules in Figure \ref{fig:NDK} make it sound and Henkin-complete
for modal higher-order logic \textbf{K}.} shown in Figure
\ref{fig:NDK}. The labels that name boxes in those inference rules are
precisely the worlds that annotate goals and hypotheses in \Coq with
the modal embedding. A hypothesis of the form \texttt{(p w)}, where
\texttt{p} is a modal proposition of type \texttt{o} and \texttt{w} is
a world of type \texttt{i} indicates that \texttt{p} is an assumption
created inside a box with name \texttt{w}.

@@

\bibliographystyle{plain}
\bibliography{Bibliography}
